Compositional Certified Resource Bounds

### 2014-06-30 paper/main.tex
typos

The precision and expressivity of the quantitative Hoare logic provide
an ample foundation for reasoning about resource consumption of Clight
programs.  The logic is naturally compositional and the derived
\emph{quantitative Hoare triples} can be used to specify the resource
consumption of library functions or to describe resource contracts
between different functions.  However, reasoning about quantitative
properties can often be more tedious then reasoning about intentional
properties.  Consequently, automation is inevitable to derive resource
bounds for large code bases like CertiKOS.  Such an automation was
easily achievable in the case of constant stack bounds for a code base
that does not contain recursive functions~\cite{veristack14}.  In
general however, resource bounds have to be parametric in the
arguments of a function and depend on the number of loop iterations
and (recursive) function calls that are performed by the function.

@@

Our \emph{second contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive as well as
\code{break} and \code{return} statements.  It targets programs that
are common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%
The use of interval sizes in potential functions (an therefore in
resource bounds) is the main innovation that makes the analysis
practical for C.

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract--interpretation base methods.  In contrast with abstract
interpretation--based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

Following the development steps of automatic amortized analysis for
functional programs~\cite{Jost03,HoffmannH10}, we deliberately
restrict our self to linear bounds in the automatic analysis (there
are no restrictions for the manually-derived bounds in the logic).
More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
where $a$ and $b$ are integer variables or constants.  The reason for
our focus on linear bounds is mainly clarity of presentation.  We
already experimented with an extension to multivariate resource
polynomials~\cite{HoffmannAH11} but this would make the inference
rules considerably more involved and should better be described
separately.  However, we developed the linear inference system so that
the extension to polynomial bounds shall work smoothly.

>>>>

The precision and expressivity of the quantitative Hoare logic provide
an ample foundation for reasoning about resource consumption of Clight
programs.  The logic is naturally compositional and the derived
\emph{quantitative Hoare triples} can be used to specify the resource
consumption of library functions or to describe resource contracts
between different functions.  However, reasoning about quantitative
properties can often be more tedious than reasoning about intentional
properties.  Consequently, automation is inevitable to derive resource
bounds for large code bases like CertiKOS.  Such an automation was
easily achievable in the case of constant stack bounds for a code base
that does not contain recursive functions~\cite{veristack14}.  In
general however, resource bounds have to be parametric in the
arguments of a function and depend on the number of loop iterations
and (recursive) function calls that are performed by the function.

@@

Our \emph{second contribution} is an automatic amortized resource
analysis for CompCert Clight that computes derivations in the
quantitative Hoare logic.  It is the first automatic amortized
resource analysis that can derive bounds that depend on negative
integers and handles programs with mutually recursive functions as well as
\code{break} and \code{return} statements.  It targets programs that
are common in system software, that is, programs whose resource
consumption can be described as a function of sizes $|[x,y]| =
\max(0,y-x)$ of intervals of integer variables.  Such bounds arise for
instance from standard \emph{for loops} \code{for (i = x; i+K \leq y;
  i=i+K)} where the step-size $K>0$ is a constant.
%
The use of interval sizes in potential functions (an therefore in
resource bounds) is the main innovation that makes the analysis
practical for C.

Despite the apparent simplicity of the new analysis system, it is able
to reproduce results from the literature (e.g,
SPEED~\cite{GulwaniMC09}) that have been obtained using sophisticated
abstract-interpretation based methods.  In contrast with abstract-interpretation
based methods, our technique does not require any
fixpoint computations to obtain loop invariants.  The mechanism we
designed is able to leverage local assertions such as $x < y$ that we
collect along the branching points of the program to obtain global
resource invariants.  We achieve this by generating a linear
constraint system that reflects resource cost and size changes of
integer variables in the program.  A solution of the linear program
immediately yields a resource usage bound for the C
program.  % We can naturally, handle
% advanced control flow such as break, return, iteration over negative
% integer variables, and mutually-recursive functions.

Following the development steps of automatic amortized analysis for
functional programs~\cite{Jost03,HoffmannH10}, we deliberately
restrict ourselves to linear bounds in the automatic analysis (there
are no restrictions for the manually-derived bounds in the logic).
More specifically, bounds have the form $\sum_{a,b} q_{(a,b)} |[a,b]|$
where $a$ and $b$ are integer variables or constants.  The reason for
our focus on linear bounds is mainly clarity of presentation.  We
already experimented with an extension to multivariate resource
polynomials~\cite{HoffmannAH11} but this would make the inference
rules considerably more involved and should better be described
separately.  However, we developed the linear inference system so that
the extension to polynomial bounds shall work smoothly.
